pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-26 article.tex
Reviewed article, progress #58, fix #151

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}], MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
allowed to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species 
evolve.
Such tools have been increasingly exploited to hypothesize and test what are the main drivers and modes for diversification processes.

@@

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the probability to realize a process given the parameters.
By maximizing it, it is possible 
to estimate what are the best values for model parameters. 
It is a standard practice in assessing such models,
to verify the level of accuracy they can achieve
on simulated phylogenies, of which full knowledge is given.

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
  }
  \label{fig:example_1}
\end{figure}

@@

The resulting figure is shown in figure 3.

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_3.png}
  \caption{
    The error BEAST2 makes from a phylogeny when
    picking the best inference model
  }
\end{figure}

@@

The resulting figure is shown in figure 4

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise.
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. 
  }
\end{figure}

@@

In figure ...

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette}
and is licensed under the GNU General Public License v3.0.
\verb;pirouette; uses the Travis CI (\url{https://travis-ci.org})
continuous integration service, which is known to significantly 
increase the number of bugs exposed (\cite{vasilescu2015}) and increases
the speed at which new features are added (\cite{vasilescu2015}).
\verb;pirouette; has a 100\% code coverage, which correlates with 
code quality (\cite{horgan1994,del1995correlation}). 
\verb;pirouette; follows Hadley Wickham's style guide (\cite{style_guide}), 
which improves software quality (\cite{fang2001}).
\verb;pirouette; depends on multiple packages, which are 
\verb;ape; (\cite{APE}), 
\verb;babette; (\cite{bilderbeek2018babette}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;mcbette; (\cite{mcbette}),
\verb;phangorn; (\cite{phangorn}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;stringr; (\cite{stringr}),
\verb;testit; (\cite{testit}) and 
\verb;usethis; (\cite{usethis}).

\verb;pirouette;'s development takes place on GitHub,
\url{https://github.com/richelbilderbeek/pirouette}, 
which accommodates collaboration (\cite{perez2016ten}) 
and improves transparency (\cite{gorgolewski2016practical}).
\verb;pirouette;'s GitHub facilitates feature requests and 
has guidelines on how to do so.

>>>>

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}], MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
allows to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species evolve.
Such tools have been increasingly exploited to hypothesize 
and test what the main drivers and modes for diversification processes are.

@@

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the probability that a phylogeny is realized 
from a given set of parameters.
Starting from a phylogeny, this likelihood is maximized
to find the parameters that are most likely.
For such diversification models, it is a standard practice 
to measure how well parameters can be recovered in this way,
by simulating 'true' phylogenies from known parameters, then
measuring the parameters deduced by maximum likelihood. 
It is a sign of quality of a diversification model 
if its parameters can be recovered well by this method.
\richel{
  We should bridge this method with how pirouette does things differently;
  why the maximum likelihood does not apply in this context  
}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
  }
  \label{fig:example_1}
\end{figure}

@@

The resulting figure is shown in figure \ref{fig:example_2}.

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny when
    picking the best inference model
  }
  \label{fig:example_2}
\end{figure}

@@

The resulting figure is shown in figure \ref{fig:example_3}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise.
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. 
  }
  \label{fig:example_3}
\end{figure}

@@

Here we discuss how to interpret the results of \verb;pirouette;.
Figure \ref{fig:example_1} shows that the generative model 
gives an error distribution with a highest density at 0.02,
where the range of errors (in this case, the nLTT statistic) is from 0.0 to 1.0.
The range of errors goes from 0.0 to 0.27. As we know the generative model,
this error distribution is the minimum noise that BEAST2 produces for this
generative model.
\richel{Show effective sample size}

Figure \ref{fig:example_2} shows that the best candidate model 
gives an error distribution with a highest density at 0.02,
where the range of errors (in this case, the nLTT statistic) is from 0.0 to 1.0.
The range of errors goes from 0.0 to 0.08. 
\richel{Describe and show evidences for the candidate model}
\richel{Describe and show the best candidate model}
This means that it is rational to use this inference model to 
infer a posterior phylogeny distribution, 
as is done by \verb;pirouette;.
\richel{Show effective sample size}
\richel{Show posterior phylogenies as densitree}

Figure \ref{fig:example_3} shows that the hand-picked model 
gives an error distribution with a highest density at 0.018,
with a range from 0.0 to 0.1.
The twin tree has an error distribution with a highest density at 0.019,
with a range from 0.0 to 0.11, which is the noise caused by the tree prior.
As there is little difference between the true and twin tree,
the (unknown) diversification model has little impact on the error
BEAST2 makes, hinting that the constant-rate birth-death model is
sufficiently good enough to explain the true phylogeny. 
The diversification model apparently has minor impact and need not to be
added to BEAST2.

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette}
and is licensed under the GNU General Public License v3.0.
\verb;pirouette; follows all practices as recommended by the
literature: continuous integration, a code coverage of 100\%
and a style guide (\cite{style_guide}).
\verb;pirouette; depends on multiple packages, which are 
\verb;ape; (\cite{APE}), 
\verb;babette; (\cite{bilderbeek2018babette}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;mcbette; (\cite{mcbette}),
\verb;phangorn; (\cite{phangorn}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;stringr; (\cite{stringr}),
\verb;testit; (\cite{testit}) and 
\verb;usethis; (\cite{usethis}).

\verb;pirouette;'s development takes place on GitHub,
\url{https://github.com/richelbilderbeek/pirouette}, 
which facilitates feature requests and 
has guidelines on how to do so.
